Bioconductor workflow for single-cell RNA-seq data analysis: dimensionality reduction, clustering, and pseudotime ordering

### 2017-06-25 workflow_combineMinSize10_mds/workflow.Rmd
more writting

---
title: "Bioconductor workflow for single-cell RNA-seq data analysis: dimensionality reduction, clustering, and pseudotime ordering"
author: 
  - name: Author Name1
    affiliation: Address of author1
  - name: Author Name2
    affiliation: Address of author2
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
bibliography: ref.bib
output: BiocWorkflowTools::f1000_article
---

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
knitr::opts_chunk$set(fig.align="center", cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
#bioconductor:
library(scRNAseq)
library(scone)
library(clusterExperiment)
library(BiocParallel)
#github
library(zinbwave)
library(slingshot)
#CRAN
library(doParallel)
library(ggplot2)
library(gplots)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(RColorBrewer)
library(digest)
library(rARPACK)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

```{r parallel}
runZinb <- TRUE
runClus <- TRUE
NCORES <- 2
mysystem = Sys.info()[['sysname']]
if (mysystem == 'Darwin'){
  registerDoParallel(NCORES)
  register(DoparParam())
}else if (mysystem == 'Linux'){
  register(bpstart(MulticoreParam(workers=NCORES)))
}else{
  print('Please change this to allow parallel computing')
  register(SerialParam())
}
```

@@

```{r frog-picture, out.width='90%', fig.align='center', fig.cap = 'Our workflow'}
knitr::include_graphics('schema_workflow.png')
```

@@

@@

The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium. To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, different lineages were assigned to each cell in the dataset and results were confirmed experimentally using in vivo lineage tracing. It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either sustentacular (support) cells or GBCs. Then, the GBC lineage, in turn, branches to give rise to olfactory sensory neurons, microvillous cells, and cells of the Bowman’s gland.

```{r stemcelldiff, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. Figure 2e from Fletcher et al paper.'}
knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
```

## Generation of the data

The dataset we use for the workflow is publicly available at https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz. It contains 849 cells and 28361 detected genes. Each cell has been labelled with labels publicly available at https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt. Pre-processing steps are as follow:
- ERCC and CreER cells were removed,
- low-quality cells were removed using package scone (ADD REF) and

FP: we should add details here.

Finally to speed up the computations, we kept only the 1000 most variable genes in this analysis. After the preprocessing steps, the dataset we are working with has 1000 genes and 747 cells. Along the workflow, we use a SummarizedExperiement object to keep the counts and the metadata in the same object

```{r datain}
if(!file.exists("../data/oe_se_1000Var.rda")) {
  source("createData.R")
}

load("../data/oe_se_1000Var.rda")
core
```

```{r batches}
batch <- colData(core)$Batch
```

Metadata for the cells is stored in colData from the SummarizedExperiment object. Cells have been processed in `r length(unique(batch))` different batches

```{r batch}
col_batch = rep(brewer.pal(9, "Set1"), 2)
names(col_batch) = unique(batch)
table(batch)
```

We also have access to the quality control measures that were used to filter low quality cells

```{r qc}
qc <- colData(core)[, !names(colData(core)) %in% c("Batch", "Experiment", "clusterLabels")]
head(qc, 2)
```

```{r clusterlabels}
clus.labels <- colData(core)[, "clusterLabels"]
```

In original work [@Fletcher2017], cells have been clustered into `r length(unique(clus.labels))` different clusters

```{r original}
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```

Note that in this dataset batches are somehow confounded with the clusters found in the original work.

@@

In single-cell RNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed. It allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest.

Here, we perform dimensionality reduction using bioconductor package zinbwave which allows to fit a zero-inflated negative binomial model (ZINB-WaVE) to get a low-dimensional representation of the data while accounting for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects.

As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use 50 dimensions and we adjust for the batches.

FP: do we want to expose the user to the equations? In the one hand, it would be easier to explain what we are looking at (W, normalized values, ...). In the other hand, it would need a good amount of explanations. I have a preference to show the equations, but what do you think?

@@

When function zinbDimRed is called, normalized values of the counts can be computed and used for visualization (e.g. in heatmaps). The normalized values are the residuals of the fit of our ZINB model where the gene and cell covariates were the same as the user specified in the zinbDimRed call but the number of unknown covariates is set to be null (i.e. K=0). When no unknown covariates are included in the model, the residuals represents the counts adjusted for the gene and cell covariates. To compute the residuals, we use the deviance residuals.

FP: note to myself: why do we have infinite values in the residuals now? It 
shows the same results as before, but we should not see infinite values here!

@@

As expected, the distributions of the normalized counts are independent of the batches

```{r lookatnorm}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- as.numeric(batch)[order(as.numeric(batch))]
boxplot(norm_order, main='Boxplot of normalized values\ncolor=batch',
        col = col_order, staplewex = 0, outline = 0, border = col_order,
        xaxt = 'n')
```

We performed principal component analysis (PCA) on the normalized values where colors are for batches on the left and original clusters on the right. As expected, there are no clustering on the left side and clustering on the right side.

```{r pcanorm}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20,
     main="PCA of normalized values\ncolor=batch")
plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20,
     main = "PCA of normalized values\ncolor=cluster")
par(mfrow = c(1,1))
```

Overall, looking at the normalized values, it seems that the batch effect was removed.

@@

Calling function zinbDimRed, we performed dimensionality reduction with K = 50. We can visualize the low dimensional matrix performing multidimensional scaling (MDS) in two dimensions. To make sure that the low dimensional matrix captured the signal of interest, the colors shown here are from the original clusters.

```{r lookatdimred}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(clus.labels)], main = 'MDS', pch = 20,
     xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

@@

We use clusterExperiment with W.

EP: I updated it to work on a SE object so that it has the meta data. If you have a summarized experiment object with W already, you could use that as long as assay(seObj) gives W.

@@

```{r examineCombineMany}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

```{r plotcoclust}
plotCoClustering(ceObj)
```

@@

FP: Elizabeth, we are working with the W here, does the locfdr make sense in this context? I set eval=FALSE in the next chunk to skip the merging step, let me know if you would rather keep using it. And if we want to still use the merging step, would we want to include it in RSEC function arguments instead of separately?

EP: I don't think the merging step on the W makes a whole lot of sense -- the method is irrelevant. The merging is based on calculating the % of genes found significant (the specific method is arbitrary). The best thing would be to replace the W with residuals in the assay of `ceObj` (or whatever data that you will do the DE on for the time stuff below), and then run the merging step on that data.  I'm not particularly fond of `locfdr`. It was probably the method that gave the best merging to Russell and Diya. You'd really have to run `mergeClusters` setting `plotInfo="all"` and look at the results and decide both the cutoff level and the method.

EP: Also, if you don't save the output of `mergeClusters` it doesn't update `ceObj`. I was calling it for just the resulting plots, since it was already merged in RSEC above. I've changed to code to update ceObj below.

FP: Ha ok, good to know. I'll keep the eval=FALSE for the moment.

```{r examineMergeClusters,eval=FALSE}
#re-does merging simpling to make plot 
#something like:
#assay(ceObj)
# if that replacement data should be considered on the transformed scale in plots, etc, the transformation function should be fixed as well:
#transformation(ceObj)
ceObj<-mergeClusters(ceObj, mergeMethod = "locfdr",
              plotInfo = "mergeMethod", cutoff = 0.01)
```

So, let's look at a heatmap on normalized values.

FP: Elizabeth, I did not find how to define the column annotation track in the plot below to have the same colors as in ceObj@clusterLegend[[1]]. I tried to use arguments annColors and annCol from aheatmap as it is said in plotHeatmap documentation that for signature matrix arguments can be passed to aheatmap. But I got the error "The following arguments to aheatmap cannot be set by the user in plotHeatmap:Rowv,Colv,color,annCol,annColors".

EP: Fanny, you would need to use the argument 'clusterLegend'. That argument takes either the format of aheatmap (list with each element a *named* vector of colors) or the format of the clusterExperiment object (i.e. list with each element a matrix with columns for `name` and `color`). So I think the following code will run, though it might need the list to have names...

But an easier fix to the code would be to set `visualizeData` option. I haven't tested this because I don't have the objects need run, so let me know if there is error.

FP: it seems great to me, what do you think?

EP: We should be careful, because the default in plotHeatmap is to plot the 500 most variable genes (maybe a slightly paternalistic default). I've changed it to `all` in the code here. I've also added the plotting of the batch, experiment, and Russell's original clusters. We may not want to keep all of them, but probably at least Russell's clusters for comparison.

```{r heatmaps}
# sampleData <- data.frame(ours = primaryCluster(ceObj))
# plotHeatmap(assays(se)$normalizedValues,
#             main = 'Normalized values, 1000 most variable genes',
#             clusterSamplesData = ceObj@dendro_samples,
#             sampleData = as.matrix(sampleData),clusterLegend=ceObj@clusterLegend[1])
# easier fix:
colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
origClusterColors<-bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
experimentColors<-bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors<-bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors<-list("Experiment"=experimentColors,"Batch"=batchColors,"clusterLabels"=origClusterColors)

plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary",clusterFeaturesData="all",
            clusterSamplesData = "dendrogramValue",
            sampleData=c("clusterLabels","Batch","Experiment"),
            clusterLegend=metaColors, annLegend=FALSE,
            main = 'Normalized values, 1000 most variable genes',
            breaks = 0.99)
```

```{r compareClusters}
plot(fit$points, col = col_clus[as.character(clus.labels)],
     main = 'MDS W, color = original clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

```{r compare}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)],
     main = 'MDS W, color = our new clusters', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = names(pal), cex = .5,
       fill = pal, title = 'Sample')
```

@@

The goal of this section is to see if we need to refit zinbwave when we want to run slingshot. We first run slingshot on the W used by clusterExperiment. In the second part of this section, we fit zinbwave on the matrix of counts where the unassigned cells have been removed. For each part (without or with refitting zinbwave), we run slingshot in the supervised and unsupervised mode and try k=3, k=4, k=5 dimensions in W.

From what I understand, start original clusters are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus). Additionally, we want the GBC cluster to be a junction before the differentiation between Microvillus and Neuron. The correspondance with the original clusters is as follow

@@

```{r kvec}
Kvec <- c(3, 4, 5)
```

### Use W components
#### Use previous W

The input of slingshot is the W used for clusterExperiment where the number of dimensions is reduced to k where k in (3, 4, 5) here.

##### Unsupervised

K = 3 only one lineage: sus is right after HBC

K = 4 same as K=3

K = 5 three lineages: hbc-gbc-immature-neuron; hbc-gbc-mv; hbc-sus (perfect!)

```{r removeunassigned}
our_cl <- primaryClusterNamed(ceObj)
cl = our_cl[our_cl != "-1"]
pal = pal[names(pal) != '-1']
```

```{r slingshot_unsup}
for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

##### Supervised

K = 3, 4, 5 get same results, all good.

```{r slingshot_sup}
for (k in Kvec){
  X <- W[our_cl != "-1", 1:k]

@@

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

#### Re-fitting zinbwave

##### Unsupervised

K = 3, 4, 5 get same results, all good.

```{r refit_zinb}
fn <- '../data/refit_zinbwave_slingshot_combineMinSize10.rda'

if (runZinb & !file.exists(fn)){
  zinbList <- lapply(Kvec, function(k){
    zinbFit(se[, our_cl != "-1"], X = '~ Batch', K = k)
  })
  save(zinbList, file = fn)
}else{
  load(fn)
}
```

```{r slingshot_unsup_refit}
for(k in Kvec) {
  X <- getW(zinbList[[k - 2]])[, 1:k]

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

##### Supervised

K = 3, 4, 5 get same results, all good.

```{r slingshot_sup_refit }
for(k in Kvec){
  X <- getW(zinbList[[k - 2]])[, 1:k]

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

### Use MDS of W

Here, instead of using the first components of W, we use the MDS of W.

#### Use previous W

The input of slingshot is the MDS of W used for clusterExperiment where the number of dimensions is k where k in (3, 4, 5) here.

##### Unsupervised

K = 3 only one lineage: sus is right after HBC

K = 4, 5 now, it is ok, three lineages: hbc-gbc-immature-neuron; hbc-gbc-mv; hbc-sus (perfect!)

```{r}
for (k in Kvec){
  X <- W[our_cl != "-1", ]
  mds <- cmdscale(dist(X), eig = TRUE, k = k)
  X <- mds$points

  lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
  curves <- get_curves(X, clus.labels = cl, lineages = lineages)
  plot_curves(X, cl, curves, col.clus = pal)
  plot_tree(X, cl, lineages, col.clus = pal)

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

##### Supervised

K = 3, 4, 5 get same results, all good.

```{r}
for (k in Kvec){
  X <- W[our_cl != "-1", ]
  mds <- cmdscale(dist(X), eig = TRUE, k = k)
  X <- mds$points

  print(paste0("K=", k))
  print(lineages$lineage1)
  print(lineages$lineage2)
  print(lineages$lineage3)
  print(lineages$lineage4)
  print(lineages$lineage5)
}
```

#### Re-fitting zinbwave

##### Unsupervised

K = 3, 4, 5 get same results, all good.

```{r}
for(k in Kvec) {
  X <- getW(zinbList[[k - 2]])
  mds <- cmdscale(dist(X), eig = TRUE, k = k)
  X <- mds$points

>>>>

---
title: "Bioconductor workflow for single-cell RNA-seq data analysis: dimensionality reduction, clustering, and pseudotime ordering"
author: 
  - name: Author Name1
    affiliation: Address of author1
  - name: Author Name2
    affiliation: Address of author2
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: single-cell, workflow, single-cell-genomics, pipeline, rna-seq, gene-expression, statistical-models, r
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of single-cell RNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
library(BiocStyle)
library(knitr)
knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)

#bioconductor:
library(scRNAseq)
library(scone)
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
library(GEOquery)
#github
library(slingshot)
#CRAN
library(doParallel)
library(ggplot2)
library(gplots)
library(magrittr)
library(matrixStats)
library(Rtsne)
library(RColorBrewer)
library(digest)
library(rARPACK)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
```

```{r parallel, echo=FALSE}
runZinb <- TRUE
runClus <- TRUE
NCORES <- 2
mysystem = Sys.info()[['sysname']]
if (mysystem == 'Darwin'){
  registerDoParallel(NCORES)
  register(DoparParam())
}else if (mysystem == 'Linux'){
  register(bpstart(MulticoreParam(workers=NCORES)))
}else{
  print('Please change this to allow parallel computing')
  register(SerialParam())
}
```

@@

```{r schema, echo=FALSE, out.width='90%', fig.cap="Workflow for analyzing single-cell RNA-seq datasets. On the right side, major plots generated by the workflow."}
knitr::include_graphics('schema_workflow.png')
```

@@

```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. This figure has been reproduced with kind permission from [@Fletcher2017].'}
knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
```

@@

The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium. To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, a single-cell RNA-seq analysis was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. Then, the expression of each gene was quantified by counting the total number of reads mapped to its cell to get a cell by gene matrix of raw counts. Different lineages were then assigned to each cell in the dataset using a set of statistical tools and results were confirmed experimentally using in vivo lineage tracing. Details on the data generation and steps of the analysis are available in [@Fletcher2017].

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman’s gland. See Figure \@ref(fig:stemcelldiff). In this study, we show possible steps to recover the lineages found in the original study from the matrix of raw counts publicly available.

FP: ask Russell if we can use the Figure \@ref(fig:stemcelldiff) here.

## Pre-processing

Counts for all genes in each cell were obtained from NCBI Gene Expression Omnibus (GEO) with accession number GSE95601. Before filtering, the dataset has 849 cells and 28361 detected genes.

```{r, echo = FALSE}
urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")

if(!file.exists("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")) {
  download.file(urls[1], "../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
  gunzip("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda.gz")
}

if(!file.exists("../data/oeHBCdiff_clusterLabels.txt")) {
  download.file(urls[2], "../data/oeHBCdiff_clusterLabels.txt")
}
```

```{r}
load("../data/GSE95601_oeHBCdiff_Cufflinks_eSet.Rda")

## count matrix
E <- assayData(Cufflinks_eSet)$counts_table

# Remove undetected genes
E <- na.omit(E)
E <- E[rowSums(E)>0,]
dim(E)
```

We removed the genes corresponding to ERCC spike-ins and the CreER gene as it corresponds to a marker gene for the HBCs cells. Davide, is that correct?

```{r}
# remove ERCC and CreER genes
cre <- E["CreER",]
ercc <- E[grep("^ERCC-", rownames(E)),]
E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
E <- E[-which(rownames(E)=="CreER"), ]
dim(E)
```

Along the workflow, we use a SummarizedExperiement object to keep the counts and the metadata in the same object. We dispose of quality control measures as well as information on the batches in which the cells were sequenced and lineages found in the original publication [@Fletcher2017]. Cells with a lineage of -2 are cells not assigned to a lineage in the orginal publication.

```{r}
# Extract QC metrics
qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))

# Extract metadata
batch <- droplevels(pData(Cufflinks_eSet)$MD_c1_run_id)
bio <- droplevels(pData(Cufflinks_eSet)$MD_expt_condition)
clusterLabels <- read.table("../data/oeHBCdiff_clusterLabels.txt",
                            sep = "\t", stringsAsFactors = FALSE)
m <- match(colnames(E), clusterLabels[, 1])

# Create metadata data.frame
metadata <- data.frame("Experiment" = bio,
                       "Batch" = batch,
                       "clusterLabels" = clusterLabels[m,2],
                       qc)

# symbol for cell not assigned to a lineage in original data
metadata$clusterLabels[is.na(metadata$clusterLabels)] <- -2

se <- SummarizedExperiment(assays = list(counts = E),
                           colData = metadata)
se
```

Using bioconductor package scone, we removed low quality cells that did not pass our quality control rule, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

```{r scone}
# Metric-based Filtering sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

mfilt <- metric_sample_filter(assay(se), nreads = colData(se)$NREADS,
                             ralign = colData(se)$RALIGN,
                             pos_controls = rownames(se) %in% hk,
                             zcut = 3, mixture = FALSE,
                             plot = TRUE)
```

```{r sconeFilt}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
se <- se[, mfilt]
dim(se)
```

Finally, to speed up the computations, we kept only the 1000 most variable genes. We recommend not to be so stringent in your analysis and perform the analysis on more genes.

```{r}
# filtering to top 1000 most variable
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
save(core, file="../data/oe_se_1000Var.rda")
```

## Dataset structure
Overall, after the preprocessing steps, our dataset has 1000 genes and 747 cells
```{r}
core
```

Metadata for the cells is stored in the object colData from the SummarizedExperiment object. Cells have been processed in 18 different batches

```{r batch}
batch <- colData(core)$Batch
col_batch = rep(brewer.pal(9, "Set1"), 2)
names(col_batch) = unique(batch)
table(batch)
```

In original work [@Fletcher2017], cells have been clustered into 14 different clusters where 151 cells were not assigned to a lineage

```{r original}
clus.labels <- colData(core)[, "clusterLabels"]
col_clus <- c("transparent", brewer.pal(12, "Set3"), brewer.pal(8, "Set2"))
col_clus <- col_clus[1:length(unique(clus.labels))]
names(col_clus) <- sort(unique(clus.labels))
table(clus.labels)
```

Note that some of the batches are confounded with the lineages found in the original work which could be a problem when we correct for batch effect in the dimensionality reduction step.

@@

In single-cell RNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analysis where clustering, pseudotime ordering, or differential expression analysis are performed. It allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low dimensional signal of interest [@Peer2017].

Here, we perform dimensionality reduction using bioconductor package zinbwave which allows to fit a zero-inflated negative binomial model (ZINB-WaVE) to get a low-dimensional representation of the data while accounting for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects. See more details in [@Risso2017].

As with most of the dimensionality reduction methods, the user needs to specify the number of dimensions of the new low dimensional space. Here, we use 50 dimensions and adjusted for the batches.

@@

When function `zinbDimRed` is called, normalized values of the counts can be computed and used for visualization (e.g. in heatmaps). The normalized values are the residuals of the fit of the ZINB-WaVE model where the gene and cell covariates were specified by the user when calling function `zinbDimRed` but no low dimensional matrix is computed. In that case, the residuals represents the counts adjusted for the gene and cell covariates and should capture the biological signal of interest. To compute the residuals, the deviance is used.

@@

As expected, the normalized values are uniformly distributed for the different batches. See Figure \@ref(fig:boxplotNorm).

```{r boxplotNorm, fig.cap="Boxplot of normalized values. Normalized values are uniformly distributed for the different batches"}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- as.numeric(batch)[order(as.numeric(batch))]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = 'n')
abline(h=0)
```

The principal component analysis (PCA) on the normalized values where colors are for batches on the left and original clusters on the right shows that, as expected, cells do not cluster by batch but by orginal published lineage. See Figure \@ref(fig:pcanorm). Overall, looking at the normalized values, it seems that the batch effect has been removed.

```{r pcanorm, fig.cap="PCA of normalized values. In the left panel, colors correspond to the different batches. In the right panel, colors correspond to the original published lineages."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = '')
plot(pca$x, col = col_clus[as.character(clus.labels)], pch = 20, main = '')
```

```{r,echo=FALSE}
par(mfrow = c(1,1))
```

@@

Calling function zinbDimRed, we performed dimensionality reduction with `K = 50` dimensions in the low dimensional space. We can visualize the low dimensional matrix (with `K = 50` dimensions) by performing multidimensional scaling (MDS) in two dimensions and euclidian distance. To make sure that the low dimensional matrix in `K = 50` dimensions captured the biological signal of interest, we plotted the MDS with colors corresponding to the original published lineages. See Figure \@ref(fig:mdsW).

```{r mdsW, fig.cap="MDS of the low dimensional matrix with K = 50 dimensions. Colors correspond to the original published lineages."}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(clus.labels)], main = '',
     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

@@

The new step of the workflow is to cluster the cells according to the low dimensional matrix computed in the previous step. We used function `RSEC` from bioconductor package clusterExperiment. Elizabeth, would you like to explain the rational and arguments of the different functions?

@@

We can visualize the resulting clusterings using the plotClusters command. See Figure \@ref(fig:examineCombineMany).

```{r examineCombineMany, fig.cap="Clusters found using function RSEC from clusterExperiment package. Cells are in the columns, and different clusterings on the rows. Each sample is color coded based on its clustering for that row, where the colors have been chosen to try to match up clusters across different clusterings that show large overlap."}
plotClusters(ceObj, colPalette = c(bigPalette, rainbow(199)))
```

We can also visualize the proportion of times these clusters were together across these clusterings. See Figure \@ref(fig:plotcoclust).

```{r plotcoclust, fig.cap="Heatmap of the proportion of times the clusters were together across the different clusterings."}
plotCoClustering(ceObj)
```

Overall, here is a summary of the number of cells clustered in each cluster

@@

```{r heatmapsClusters, fig.cap="Heatmap of the normalized values for the 1000 most variable genes (rows)."}
# set colors for cell clusterings
colData(ceObj)$clusterLabels <- as.factor(colData(ceObj)$clusterLabels)
origClusterColors<-bigPalette[1:nlevels(colData(ceObj)$clusterLabels)]
experimentColors<-bigPalette[1:nlevels(colData(ceObj)$Experiment)]
batchColors<-bigPalette[1:nlevels(colData(ceObj)$Batch)]
metaColors<-list("Experiment"=experimentColors,"Batch"=batchColors,"clusterLabels"=origClusterColors)

plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary",clusterFeaturesData="all",
            clusterSamplesData = "dendrogramValue",
            sampleData=c("clusterLabels","Batch","Experiment"),
            clusterLegend=metaColors, annLegend=FALSE,
            main = '',
            breaks = 0.99)
```

We can now plot the MDS of the low dimensional matrix coloring the cells by their newly found clusters. See Figure \@ref(fig:mdsWce) which is the same as Figure \@ref(fig:mdsW) but where colors for the cells are different.

```{r mdsWce, fig.cap="MDS of the low dimensional matrix with K = 50 dimensions. Colors correspond to the clusters found using our workflow."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = '', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = names(pal), cex = .5,
       fill = pal, title = 'Sample')
```

@@

We now want to order the clusters found using function `RSEC` from package clusterExperiment. From original published work, we know that the start clusters are 1 and 5 (HBC) and end original clusters are 15 (Microvillus), 9 and 12 (neuron), and 4, 7 (Sus). Additionally, we know that the GBC cluster should be a junction before the differentiation between Microvillus cells and Neurons. The correspondance with the original clusters is as follow

@@

As the input for the speudotime ordering algorithm, we use the MDS of low dimensional matrix (K=50) where the number of dimensions for MDS is `k=3`. The algorithm is stable to the number of dimensions k in the MDS. We use the unsupervised version of slingshot where we only give the information of the start cluster but not the end clusters.

Kelly, would you like to give more information about slingshot?

```{r curves}
X <- W[our_cl != "-1", ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

lineages <- get_lineages(X, clus.labels = cl, start.clus = "c1")
curves <- get_curves(X, clus.labels = cl, lineages = lineages)
plot_curves(X, cl, curves, col.clus = pal)
```

```{r tree}
plot_tree(X, cl, lineages, col.clus = pal)
```

```{r lineages}
print(lineages$lineage1)
print(lineages$lineage2)
print(lineages$lineage3)
```

In this workflow, it is ok to use unsupervised. In some other cases, we have noticed that we need to give more knowledge to the algorithm to find the correct ordering. Here is the code to use slingshot in a supervised setting.

```{r slingshotsupervised, eval=FALSE}
X <- W[our_cl != "-1", ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

@@

print(lineages$lineage1)
print(lineages$lineage2)
print(lineages$lineage3)
print(lineages$lineage4)
print(lineages$lineage5)
```
